Supporting Information: Improving cropland maps through tight integration of human and machine intelligence

### 2020-08-07 inst/manuscript/figures.Rmd
Updated analyses and SI, manuscript text to include results from active versus random training

```{r setup, include=FALSE, cache=FALSE, message = FALSE}
library(raster)
library(sf)
library(dplyr)
library(ggplot2)
library(here)
library(RColorBrewer)
library(activemapper)
```

@@

```{r, eval=FALSE}
pth <- function(x) file.path(here("inst/manuscript"), x)

@@

# Plots
cols <- c(rep("grey", length(unique(iteration_metrics$AOI)) -1), "black")
lsize <- c(rep(0.3, length(unique(iteration_metrics$AOI)) -1), 1)
V1 <- c("Accuracy", "AUC", "F1")

# prepare data
# remove labeller from AOI
imetrics <- iteration_metrics %>% 
  filter(Metric %in% V1) %>% 
  mutate(AOI = gsub("labeller", "", AOI))

@@

# plot
p <- ggplot(imetrics) + 
  geom_line(aes(Iteration, Score, color = AOI, size = AOI)) +
  geom_point(data =  mx_metrics, aes(Iteration, Score, color = AOI), 
             size = 1) +
  ggrepel::geom_text_repel(
    data = mx_metrics %>% filter(AOI != "All"), 
    aes(label = AOI, x = Iteration, y = Score), col = "grey50", 
    segment.color = "grey80", point.padding = 0.05 
  ) +
  scale_color_manual(values = cols, guide = FALSE) +
  scale_size_manual(values = lsize, guide = FALSE) + 
  scale_y_continuous(breaks = seq(0.2, 1, 0.1), expand = c(0, 0), 
                     limits = c(0.2, 1)) + 
  facet_wrap(vars(Metric), ncol = 3) + 
  theme_linedraw() + 
  theme(strip.background = element_rect(fill = "white"), 
        theme(strip.text = element_text(colour = "black"))

@@

### Figure S6: Consenus label (Bayesian) risk

@@

```{r, eval=FALSE}
# randomly select cells, as if those used to assess image quality
set.seed(1)
tiles_ss <- tiles %>% sample_n(100)

# set up dummy values of quality, using rainfall as example and regression with
# some random noise added
rftiles <- unlist(extract(rftot, tiles_ss))  # rainfall
tiles_ss <- tiles_ss %>% mutate(rf = rftiles) %>% select(aoi, tile, rf)
rflm <- coef(lm(c(8, 1) ~ range(tiles_ss$rf)))  # regression 1 to 8
qual_preds <- rflm[1] + rflm[2] * tiles_ss$rf  # predict
set.seed(7) 
qual_preds <- round(qual_preds *  # random noise
                      runif(length(qual_preds), min = 0.7, max = 1.3))
qual_preds <- ifelse(qual_preds > 5, 5, qual_preds)  # force to 5
tiles_ss <- tiles_ss %>% mutate(Quality = factor(qual_preds)) %>% 
  select(aoi, tile, rf, Quality)

# random assignment of 1-5 by rainfall
p1 <- ggplot(africa) + geom_sf(fill = "gray97", color = "grey", size = 0.25) +
  geom_sf(data = tiles, fill = "grey85", col = "grey", size = 0.3) +
  geom_sf(data = ghana, fill = "transparent", col = "grey", size = 0.3) +
  geom_sf(data = tiles_ss, 
          mapping = aes(fill = Quality, color = Quality)) +   
  scale_fill_brewer(palette = "RdBu") + scale_color_brewer(palette = "RdBu") + 
  coord_sf(xlim = maprange[1:2], ylim = maprange[3:4]) + 
  theme_void()
p2 <- ggplot(tiles_ss) + 
  geom_histogram(aes(x = Quality), bins = 5, stat = "count", col = "grey", 
                 fill = "blue") + 
  xlab("Quality of composite") + ylab("N tiles") + 
  cowplot::theme_cowplot(font_size = 12)
g <- cowplot::plot_grid(p1, p1, p2, p2, nrow = 2, scale = c(1, 1, 0.9), 
                        labels = LETTERS[1:4], label_x = 0.05)
ggsave("manuscript/figures/figure4.png", width = 5, height = 5, 
       units = "in", dpi = 300)
```

### Figure S7: Segmentation statistics
```{r segstats, eval=FALSE}
data("top_label_data")
data("field_validation_stats")

>>>>

```{r setup, include=FALSE, cache=FALSE, message = FALSE}
library(raster)
library(sf)
library(dplyr)
library(ggplot2)
library(here)
library(RColorBrewer)
library(activemapper)
pth <- function(x) file.path(here("inst/manuscript"), x)
```

@@

```{r, eval=FALSE}

@@

# Plots
V1 <- c("Accuracy", "AUC", "F1")

# prepare data
# remove labeller from AOI and select AOIs 1-16
aois <- c(paste0("labeller", 1:16), "All")
imetrics <- iteration_metrics %>% 
  filter(Metric %in% V1) %>% 
  filter(AOI %in% aois) %>% 
  mutate(AOI = gsub("labeller", "", AOI))

@@

# plot
cols <- c(rep("grey", length(unique(imetrics$AOI)) -1), "black")
lsize <- c(rep(0.3, length(unique(imetrics$AOI)) -1), 1)
p <- ggplot(imetrics) + 
  geom_line(aes(Iteration, Score, color = AOI, size = AOI)) +
  geom_point(data = mx_metrics, aes(Iteration, Score, color = AOI), 
             size = 1.2) +
  ggrepel::geom_text_repel(
    data = mx_metrics %>% filter(AOI != "All"), 
    aes(label = AOI, x = Iteration, y = Score), col = "grey50", 
    segment.color = "grey60", point.padding = 0.1, min.segment.length = 0.2,
    segment.size = 0.2, size = 3   
  ) +
  scale_color_manual(values = cols, guide = FALSE) +
  scale_size_manual(values = lsize, guide = FALSE) + 
  scale_x_continuous(expand = c(0.01, 0.01)) + 
  scale_y_continuous(breaks = seq(0.2, 1, 0.1), expand = c(0.01, 0.01), 
                     limits = c(0.2, 1)) + 
  facet_wrap(vars(Metric), ncol = 3) + 
  theme_linedraw() + 
  theme(strip.background = element_rect(fill = "white"), 
        strip.text = element_text(colour = "black"))

@@

### Figure S6: Consensus label (Bayesian) risk

@@

### Figure S7: Random versus active sampling
```{r, eval=FALSE}
data("iteration_metrics")

# Plots
V1 <- c("Accuracy", "AUC", "F1")

# prepare data
# remove labeller from AOI and select AOIs 1-16
aois <- c(paste0("labeller", c(1, 8, 15, "1r", "8r", "15r")))
imetrics <- iteration_metrics %>% 
  filter(Metric %in% V1 & AOI %in% aois) %>% 
  mutate(AOI = gsub("labeller", "", AOI)) %>% 
  mutate(type = case_when(
    AOI %in% c(1, 8, 15) ~ "active", 
    AOI %in% c("1r", "8r", "15r") ~ "random"
  )) %>% 
  mutate(AOI = gsub("r", "", AOI)) %>% 
  tidyr::pivot_wider(names_from = "type", values_from = "Score") %>% 
  mutate(delta = (active - random) / random * 100)

imetrics_all <- imetrics %>% filter(Iteration < 4) %>% 
  group_by(Iteration, Metric) %>% 
  summarize(active = mean(active), random = mean(random), 
            delta = mean(delta)) %>%
  ungroup() %>% mutate(AOI = "All") %>% 
  select(AOI, Iteration, Metric, active, random, delta)
imetrics <- bind_rows(imetrics, imetrics_all)

# plot
cols <- c(rep("grey50", length(unique(imetrics$AOI)) -1), "black")
lsize <- c(rep(0.3, length(unique(imetrics$AOI)) -1), 1)
p <- ggplot(imetrics) + 
  geom_hline(yintercept = 0, linetype = "dotted", col = "grey40") +
  geom_line(aes(Iteration, delta, color = AOI, size = AOI)) +
  geom_point(data = mx_metrics, aes(Iteration, delta, color = AOI), 
             size = 1.2) +
  ggrepel::geom_text_repel(
    data = mx_metrics %>% filter(AOI != "All"), 
    aes(label = AOI, x = Iteration, y = delta), col = "grey50", 
    segment.color = "grey60", point.padding = 0.2, min.segment.length = 0.2,
    segment.size = 0.2, size = 3, nudge_x = 0 
  ) +
  scale_color_manual(values = cols, guide = FALSE) +
  scale_size_manual(values = lsize, guide = FALSE) + 
  scale_x_continuous(expand = c(0.01, 0.01)) + 
  scale_y_continuous(expand = c(0.01, 0.01), breaks = seq(-8, 4, 1)) + 
  ylab("% difference") +                    
  facet_wrap(vars(Metric), ncol = 3) +
  theme_linedraw() + 
  theme(panel.grid.minor.x = element_blank(), 
        strip.background = element_rect(fill = "white"), 
        strip.text = element_text(colour = "black"))

ggsave(pth("figures/si_random_vs_active.png"), 
       width = 6, height = 4, units = "in", dpi = 300)

### Figure S8: Segmentation statistics
```{r segstats, eval=FALSE}
data("top_label_data")
data("field_validation_stats")

### 2020-08-07 inst/manuscript/manuscript.Rmd
Updated analyses and SI, manuscript text to include results from active versus random training

- name: Azavea
  affilnum: '7'

@@

- affilnum: 5
  affil: Meridia

- affilnum: 5
  affil: Azavea

@@

### Map assessment
#### Validation performance
To assess the gain in RandomForest performance due to active learning loop, we assessed the change in accuracy, F1, and AUC (see 2.2.3.2) between each iteration and between the first and last iterations for each AOI. We then compared the magnitude of these iteration-wise changes to those that occurred when each batch of new training sites was randomly selected [@DebatsIntegratingactivelearning2017].

@@

@@

# differences per iteration
imet_dmu <- iteration_metrics %>% 
  filter(AOI == "All" & Iteration > 0) %>%
  filter(grepl("_change", Metric)) %>% 
  mutate(Metric = gsub("_change", "", Metric)) %>% 
  mutate(Score = round(Score * 100, 1)) %>% 
  filter(Metric %in% c("Accuracy", "AUC", "F1"))

@@

# AOI score ranges
imet_min <- iteration_metrics %>% group_by(AOI) %>%
  filter(Iteration == max(Iteration) & !grepl("change", Metric)) %>%
  filter(Metric %in% c("Accuracy", "AUC", "F1")) %>% 
  ungroup() %>% 
  group_by(Metric) %>% 
  filter(Score == min(Score)) %>% 
  mutate(AOI = gsub("labeller", "", AOI), Score = round(Score, 3))

imet_max <- iteration_metrics %>% group_by(AOI) %>%
  filter(Iteration == max(Iteration) & !grepl("change", Metric)) %>%
  filter(Metric %in% c("Accuracy", "AUC", "F1")) %>% 
  ungroup() %>% 
  group_by(Metric) %>% 
  filter(Score == max(Score)) %>% 
  mutate(AOI = gsub("labeller", "", AOI), Score = round(Score, 3))
```

The gain in model performance was assessed for each AOI against several accuracy metrics. The average accuracy, AUC, and F1 at Iteration 0 were `r imet_mu0 %>% filter(Metric == "Accuracy") %>% pull(Score)`, `r imet_mu0 %>% filter(Metric == "AUC") %>% pull(Score)`, and `r imet_mu0 %>% filter(Metric == "F1") %>% pull(Score)`, respectively, increasing to `r imet_mu3 %>% filter(Metric == "Accuracy") %>% pull(Score)`, `r imet_mu3 %>% filter(Metric == "AUC") %>% pull(Score)`, and `r imet_mu3 %>% filter(Metric == "F1") %>% pull(Score)` by Iteration 3 (Figure \ref{fig:alperformance}). These differences represent respective gains of `r imet_mu3 %>% filter(Metric == "Accuracy") %>% pull(dscore)`, `r imet_mu3 %>% filter(Metric == "AUC") %>% pull(dscore)`, and
`r imet_mu3 %>% filter(Metric == "F1") %>% pull(dscore)` for the three metrics. The largest gains in each metric were registered on Iteration 1, averaging `r imet_dmu1 %>% filter(Metric == "Accuracy") %>% pull(Score)`, `r imet_dmu1 %>% filter(Metric == "AUC") %>% pull(Score)`, and `r imet_dmu1 %>% filter(Metric == "F1") %>% pull(Score)` percent for accuracy, AUC, and F1. The lowest gains were received on Iteration 3 for accuracy (`r imet_dmu3 %>% filter(Metric == "Accuracy") %>% pull(Score)`%) and F1 (`r imet_dmu3 %>% filter(Metric == "F1") %>% pull(Score)`%), while Iteration 2 showed the lowest average gain for AUC (`r imet_dmu3 %>% filter(Metric == "AUC") %>% pull(Score)`%). The range in scores varied substantially across AOIs and metrics. Accuracy ranged between `r imet_min %>% filter(Metric == "Accuracy") %>% pull(Score)` (AOI `r imet_min %>% filter(Metric == "Accuracy") %>% pull(AOI)`) and `r imet_max %>% filter(Metric == "Accuracy") %>% pull(Score)` (AOI `r imet_max %>% filter(Metric == "Accuracy") %>% pull(AOI)`), while AUC varied from `r imet_min %>% filter(Metric == "AUC") %>% pull(Score)` (AOI `r imet_min %>% filter(Metric == "AUC") %>% pull(AOI)`) and `r imet_max %>% filter(Metric == "AUC") %>% pull(Score)` (AOI `r imet_max %>% filter(Metric == "AUC") %>% pull(AOI)`), and F1 from `r imet_min %>% filter(Metric == "F1") %>% pull(Score)` (AOI `r imet_min %>% filter(Metric == "F1") %>% pull(AOI)`) and `r imet_max %>% filter(Metric == "F1") %>% pull(Score)` (AOI `r imet_max %>% filter(Metric == "F1") %>% pull(AOI)`).

@@

We compared the gain in performance at AOIs 1, 8, and 15 against models trained for those same locations using the initial random draw of 500, followed by increments of 100 randomly selected points per iteration.

>>>>

@@

- affilnum: 6
  affil: Meridia

@@

### Map assessment
#### Validation performance
To assess the gain in RandomForest performance due to active learning, we assessed the change in accuracy, F1, and AUC (see 2.2.3.2) between each iteration and between the first and last iterations for each AOI. To assess whether active learning improved the performance relative to a purely random approach to selecting new training sites, we ran an additional test at a subset of AOIs (1, 8, and 15). We first randomly selected and labelled 300-400 sites in each AOI. After initial training of AOIs 1, 8, and 15 with their initial pools of 500 training sites (Cluster 1 for AOI 1 and Cluster 2 for AOIs 8 and 15), we then retrained the model for each AOI by progressively adding 100 of the new randomly selected labels, evaluating model performance against the validation set after each retraining. We repeated this process so that the number of retrainings and training samples matched those from the active learning process for each AOI. We then compared the difference in accuracy, AUC, and F1 between these randomly trained models and those trained with active learning [@DebatsIntegratingactivelearning2017].

@@

V1 <- c("Accuracy", "AUC", "F1")

@@

# differences per iteration
imet_dmu <- iteration_metrics %>% 
  filter(AOI == "All" & Iteration > 0) %>%
  filter(grepl("_change", Metric)) %>% 
  mutate(Metric = gsub("_change", "", Metric)) %>% 
  mutate(Score = round(Score * 100, 1)) %>% 
  filter(Metric %in% V1)

@@

# AOI score ranges
imin <- iteration_metrics %>% 
  filter(AOI %in% paste0("labeller", 1:16)) %>% group_by(AOI) %>%
  filter(Iteration == max(Iteration) & !grepl("change", Metric)) %>%
  filter(Metric %in% V1) %>% 
  ungroup() %>% 
  group_by(Metric) %>% 
  filter(Score == min(Score)) %>% 
  mutate(AOI = gsub("labeller", "", AOI), Score = round(Score, 3))

imax <- iteration_metrics %>% 
  filter(AOI %in% paste0("labeller", 1:16)) %>% group_by(AOI) %>%
  filter(Iteration == max(Iteration) & !grepl("change", Metric)) %>%
  filter(Metric %in% V1) %>% 
  ungroup() %>% 
  group_by(Metric) %>% 
  filter(Score == max(Score)) %>% 
  mutate(AOI = gsub("labeller", "", AOI), Score = round(Score, 3))

# summarize and condense for easier in line extraction
# iteration 0 and 0-3 metrics
imu0 <- lapply(V1, function(x) {
  imu <- round(imet_mu0 %>% filter(Metric == x) %>% pull(Score), 3)
  idf <- round(imet_mu0 %>% filter(Metric == x) %>% pull(dscore), 3)
  ix <- list("score" = imu, "diff" = idf)
  ix
})
names(imu0) <- V1

imu3 <- lapply(V1, function(x) {
  imu <- round(imet_mu3 %>% filter(Metric == x) %>% pull(Score), 3)
  idf <- round(imet_mu3 %>% filter(Metric == x) %>% pull(dscore), 3)
  ix <- list("score" = imu, "diff" = idf)
  ix
})
names(imu3) <- V1

# iteration-wise diffs
idmu <- lapply(list(imet_dmu1, imet_dmu2, imet_dmu3), function(x) {
  l <- lapply(V1, function(y) {
    imu <- round(x %>% filter(Metric == y) %>% pull(Score), 3)
    idf <- round(x %>% filter(Metric == y) %>% pull(dscore), 3)
    ix <- list("score" = imu, "diff" = idf)
    ix
  })
  names(l) <- V1
  l
})
names(idmu) <- paste0("I", 1:3)
```

The gain in model performance was assessed for each AOI against several accuracy metrics. The average accuracy, AUC, and F1 at iteration 0 were `r imu0$Accuracy$score`, `r imu0$AUC$score`, and `r imu0$F1$score`, respectively, increasing to `r imu3$Accuracy$score`, `r imu3$AUC$score`, and `r imu3$F1$score` by iteration 3 (Figure \ref{fig:alperformance}). These differences represent respective gains of `r imu3$Accuracy$diff`, `r imu3$AUC$diff`, and `r imu3$F1$diff` percent for the three metrics. The largest gains in each metric were registered on iteration 1, averaging `r idmu$I1$Accuracy$score`, `r idmu$I1$AUC$score`, and `r idmu$I1$F1$score` percent for accuracy, AUC, and F1. The lowest gains were received on Iteration 3 for accuracy (`r idmu$I3$Accuracy$score`%) and F1 (`r idmu$I3$F1$score`%), while AUC gains were least on iteration 2 (`r idmu$I2$AUC$score`%). The range in scores varied substantially across AOIs and metrics. Accuracy ranged between `r imin[imin$Metric == "Accuracy", ]$Score` (AOI `r imin[imin$Metric == "Accuracy", ]$AOI`) and `r imax[imax$Metric == "Accuracy", ]$Score` (AOI `r imax[imax$Metric == "Accuracy", ]$AOI`), while AUC varied from `r imin[imin$Metric == "AUC", ]$Score` (AOI `r imin[imin$Metric == "AUC", ]$AOI`) and `r imax[imax$Metric == "AUC", ]$Score` (AOI `r imax[imax$Metric == "AUC", ]$AOI`), and F1 from `r imin[imin$Metric == "F1", ]$Score` (AOI `r imin[imin$Metric == "F1", ]$AOI`) and `r imax[imax$Metric == "F1", ]$Score` (AOI `r imax[imax$Metric == "F1", ]$AOI`).

@@

```{r, echo=FALSE}
# random versus active data
data("iteration_metrics")

# process data: remove labeller from AOI and select AOIs 1-16
V1 <- c("Accuracy", "AUC", "F1")
aois <- c(paste0("labeller", c(1, 8, 15, "1r", "8r", "15r")))
imetrics <- iteration_metrics %>% 
  filter(Metric %in% V1 & AOI %in% aois) %>% 
  mutate(AOI = gsub("labeller", "", AOI)) %>% 
  mutate(type = case_when(
    AOI %in% c(1, 8, 15) ~ "active", 
    AOI %in% c("1r", "8r", "15r") ~ "random"
  )) %>% 
  mutate(AOI = gsub("r", "", AOI)) %>% 
  tidyr::pivot_wider(names_from = "type", values_from = "Score") %>% 
  mutate(delta = (active - random) / random * 100) 
imetrics_all <- imetrics %>% 
  filter(Iteration < 4) %>% 
  group_by(Iteration, Metric) %>% 
  summarize(active = mean(active), random = mean(random), 
            delta = mean(delta)) %>%
  ungroup() %>% mutate(AOI = "All") %>% 
  select(AOI, Iteration, Metric, active, random, delta)

ixl <- lapply(1:3, function(x) {
  ix <- imetrics_all %>% filter(Iteration == x) %>% pull(delta)
  ix <- round(ix, 1)
  names(ix) <- V1
  ix
})
names(ixl) <- as.roman(1:3)

i1acc <- imetrics %>% 
  filter(Iteration == 1 & Metric == "Accuracy") %>% 
  pull(delta) %>% round(2)
names(i1acc) <- imetrics %>% distinct(AOI) %>% pull()

u <- function(x) unname(x)
```

The training tests at AOIs 1, 8, and 15 showed that active learning produced higher performance than randomized model training after three iterations, with accuracy, AUC, and F1 being respectively `r u(ixl$III["Accuracy"])`, `r u(ixl$III["AUC"])`, and `r u(ixl$III["F1"])` percent higher after three iterations (Figure S5). However, there was more variability in earlier iterations, with average score differences of `r u(ixl$I["Accuracy"])` (accuracy), `r u(ixl$I["AUC"])` (AUC), and `r u(ixl$I["F1"])` percent (F1) after iteration 1, and `r u(ixl$II["Accuracy"])` (accuracy), `r u(ixl$II["AUC"])` (AUC), and `r u(ixl$II["F1"])` (F1) percent after iteration 2. The lower accuracy under active leaving was caused by results at AOI 15, where accuracy gain was `r abs(u(i1acc["15"]))` percent lower under active training (although accuracy slightly between iterations; see Figure \ref{fig:alperformance}). In comparison, active learning accuracies after iteration 1 were `r u(i1acc["1"])` and `r u(i1acc["8"])` percent higher than random training for AOIs 1 and 8, respectively.

### 2020-08-07 inst/manuscript/si.Rmd
Updated analyses and SI, manuscript text to include results from active versus random training

>>>>

### Validation
The differences in accuracy, AUC, and F1 between the active learning process and the random retraining at each iteration for AOIs 1, 8, and 15 (Figure \ref{fig:randomvactive}). Small differences due to random variations in the RandomForest models are the reason for the non-zero differences at iteration 0, when both models were trained with the same level set.   
```{r randomvactive, echo = FALSE, out.width="90%", fig.cap="The percent difference in performance metrics per iteration for AOIs 1, 8, and 15 (grey lines with numbers indicating AOI; the black line indicates average difference across the three AOIs) when comparing models trained using active learning versus those trained using randomly selected sites. Postive percentages indicate superior performance by active learning, negative percentages the inverse.", fig.align='center', message=FALSE, results='hold'}
knitr::include_graphics('figures/si_random_vs_active.png')
```
